---
title: "Adapting Algorithms for CAISEr"
author: "Felipe Campelo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adapting Algorithms for CAISEr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
This is a short guide to adapting existing algorithms and problem instances for 
running an experiment using CAISEr. In this document, we cover:

- Definition of instance lists
- Adaptation of existing algorithms
- Some stuff you can do with the results

A general description of the CAISE methodology is available in our papers:
- F. Campelo, F. Takahashi, _Sample size estimation for power and accuracy in the  experimental comparison of algorithms_. Journal of Heuristics 25(2):305-338, 
2019.]  
- F. Campelo, E. Wanner, _Sample size calculations for the experimental comparison of multiple algorithms on multiple problem instances_. Submitted, Journal of 
Heuristics, 2019.

## Assembling an instance list
As stated in the documentation of both `run_experiment()` and `calc_nreps()`, each 
instance must be a named list containing all relevant parameters that 
define the problem instance. This list must contain at least the field
`instance$FUN`, with the name of the problem instance function, that is, a
routine that calculates $y = f(x)$. If the instance requires additional
parameters, these must also be provided as named fields. Each instance can 
also have an `alias`, a unique name to distinguish it from other instances. If 
no alias is provided, the name of the function (`instance$FUN`) is used as the 
instance ID.

The `instances` parameter for `run_experiment()` is simply a list or vector of 
these instance lists.

To illustrate how to adapt existing implementations to this structure, we 
assume that we are interested in comparing three multiobjective optimization 
algorithms for a (hypothetical) problem class represented by 
problems UF1 - UF7 (in dimensions between 10 and 40) from package 
[smoof](https://CRAN.R-project.org/package=smoof). For this implementation to 
work with the `MOEADr::moead()` routine (see next section) some 
manipulation is necessary, but the instance list in this case is simply a list 
with each element containing the name of the routine as field `$FUN` (since 
all function names are different, no need for aliases).

```{r}
suppressPackageStartupMessages(library(smoof))
suppressPackageStartupMessages(library(MOEADr))

### Build function names (instances: UF1 - UF7, dimensions 10 - 40)
fname   <- paste0("UF_", 1:7)
dims    <- c(10:40)
allfuns <- expand.grid(fname, dims, stringsAsFactors = FALSE)

# Assemble instances list
instances <- vector(nrow(allfuns), mode = "list")
for (i in 1:length(Instance.list)){
  instances[[i]]$FUN <- paste0(allfuns[i,1], "_", allfuns[i,2])
}

### Build the functions listed above (so that they can be properly used)
for (i in 1:nrow(allfuns)){
  assign(x = instances[[i]]$FUN,
     value = MOEADr::make_vectorized_smoof(prob.name  = "UF",
                    dimensions = allfuns[i, 2],
                    id = as.numeric(strsplit(allfuns[i, 1], "_")[[1]][2])))
}
```

## Adaptation of an existing algorithm implementation
We will use the MOEA/D implementation available in the [MOEADr](https://CRAN.R-project.org/package=MOEADr) package as our base 
algorithm, and assume that we are interested in comparing the performance of 
three versions of this algorithm: the two versions described in the [original MOEA/D paper](https://ieeexplore.ieee.org/document/4358754), and the one described in the [MOEA/D-DE paper](https://ieeexplore.ieee.org/document/4633340) (see the documentation of [MOEADr](https://CRAN.R-project.org/package=MOEADr) and 
references therein for details of these methods) as solvers of the hypothetical 
problem class represented by the available test instances. The performance of 
each algorithm on each instance will be measured according to an indicator 
known as _Inverted Generational Distance_ (IGD - details [here](http://ieeexplore.ieee.org/document/1197687/)), for which 
_smaller = better_.

As described in the documentation of both `run_experiment()` and 
`calc_nreps()`, an `algorithm` must contain an `algorithm$FUN` field (the name
of the function that calls the algorithm) and any other elements/parameters
that `algorithm$FUN` requires (e.g., stop criteria, operator names and
parameters, etc.). An additional field, `algorithm$alias`, can be used to
provide the algorithm with a unique identifier. 

Supposing that the list in `algorithm` has fields `algorithm$FUN = myalgo`, `algorithm$par1 = "a"`, `algorithm$par2 = 5`, then the function in 
`algorithm$FUN` must have the following structure:

```
myalgo <- function(par1, par2, instance, ...){
  # do stuff
  return(results)
}
```

That is, it must be able to run if called as:

```
# remove '$FUN' and '$alias' field from list of arguments
# and include the problem definition as field 'instance'
myargs          <- algorithm[names(algorithm) != "FUN"]
myargs          <- myargs[names(myargs) != "alias"]
myargs$instance <- instance

# call 'algorithm$FUN' with the arguments in 'myargs'
do.call(algorithm$FUN, args = myargs)
```

Finally, the `algorithm$FUN` routine must return a list object containing (at
least) the performance value of the final solution obtained after a given
run, in a field named `value` (e.g., `result$value`) .

To build the algorithm functions to be used in `run_experiment()`, we 
encapsulate (almost) all algorithm parameters within a `myalgo()` function, 
which receives only two inputs: the instance to be solved (i.e., one element 
from `Instance.list`) and the specification of which version of the algorithm 
is to be run.

```{r}
# Prepare algorithm function to be used in run_experiment():
myalgo <- function(type, instance){
  # Input parameters:
  #     - type (variant to use: "original", "original2" or "moead.de")
  #     - instance (instance to be solved, e.g., instance = instances[[i]])
  # All other parameters are set internally

  ## Extract instance information to build the MOEADr problem format
  fdef  <- unlist(strsplit(instance$FUN, split = "_"))
  uffun <- smoof::makeUFFunction(dimensions = as.numeric(fdef[3]),
                                 id         = as.numeric(fdef[2]))
  fattr    <- attr(uffun, "par.set")
  prob.dim <- fattr$pars$x$len
  
  ## Build MOEADr problem list
  problem <- list(name = instance$FUN,
                  xmin = fattr$pars$x$lower,
                  xmax = fattr$pars$x$upper,
                  m    = attr(uffun, "n.objectives"))

  ## Load presets for the algorithm provided in input 'type' and 
  ## modify whatever is needed for this particular experiment
  
  algo.preset <- MOEADr::preset_moead(type)
  algo.preset$decomp$H <- 99 # <-- set population size
  algo.preset$stopcrit[[1]]$name <- "maxeval" # <-- type of stop criterion
  algo.preset$stopcrit[[1]]$maxeval <- 2000 * prob.dim # stop crit.
  poly.ind <- which(sapply(algo.preset$variation,
                           function(x){x$name == "polymut"}))
  algo.preset$variation[[poly.ind]]$pm <- 1 / prob.dim # <--- pm = 1/d
  
  ## Run algorithm on "instance"
  out <- MOEADr::moead(preset = algo.preset, problem = problem,
                       showpars = list(show.iters = "none"))

  ## Read reference data to calculate the IGD
  Yref  <- as.matrix(read.table(paste0("../inst/extdata/pf_data/",
                                       fdef[1], fdef[2], ".dat")))
  IGD = MOEADr::calcIGD(Y = out$Y, Yref = Yref)

  ## Return IGD as field "value" in the output list
  return(list(value = IGD))
}
```

Finally, the `algorithms` parameter must be assembled as a list of 
algorithm objects (each containing fields `$FUN`, `$alias` and, in 
this case, `$type`).

```{r}
# Assemble Algorithm.list. Notice that we need to provide an alias for each 
# method, since both algorithms have the same '$FUN' argument.
algorithms <- list(list(FUN   = "myalgo", 
                        alias = "Original 1", 
                        type  = "original"),
                   list(FUN   = "myalgo", 
                        alias = "Original 2", 
                        type  = "original2"),
                   list(FUN   = "myalgo", 
                        alias = "MOEAD-DE", 
                        type  = "moead.de"))
```

## Running an experiment using CAISEr

With the definitions above it is possible now to run an experiment 
using the iterative sample size determination implemented in CAISEr. For 
that, all we have to do is define the desired experimental parameters and 
use `run_experiment()`:

```{r, eval=FALSE}
library(CAISEr)
my.results <- run_experiment(instances  = instances,
                             algorithms = algorithms,
                             power = 0.8,      # Desired power: 80%
                             power.target = "mean", # on average,
                             d = 0.5,          # to detect differences greater
                                               # than 0.5 standard deviations
                             sig.level = 0.05, # at a 95% confidence level. 
                             se.max = 0.05,    # Measurement error: 5% 
                             dif = "perc",     # on the paired percent 
                                               # differences of means,
                             method = "param", # calculated using parametric
                                               # formula.
                             comparisons = "all.vs.all", # Compare all algorithms 
                                                         # vs all others,
                             nstart = 15,      # start with 15 runs/algo/inst
                             nmax   = 200,     # and do no more than 200 runs/inst.
                             seed   = 1234,    # PRNG seed (for reproducibility)
                             ncpus  = parallel::detectCores() - 1)
```

```{r, echo=FALSE}
load("../inst/extdata/vignette_results.RData")
```

After that we can interrogate the results and perform inference, if we are so 
inclined. 
For instance, we can check if our sample of paired differences in performance 
is (at least approximately) Normal, so that we can assume a Normal sampling 
distribution of the means and use a t test with a clean conscience:

```{r, fig.align="center", fig.width=7, fig.height=5}
# Take a look at the data summary:
summary(my.results)
# print(my.results)

suppressPackageStartupMessages(library(car))
car::qqPlot(my.results$data.summary$phi.j, 
            pch = 20, las = 1, 
            ylab = "observed results", xlab = "theoretical quantiles")
```

The normal QQ plot indicates that no expressive deviations of normality are 
present, which gives us confidence in using the t test as our inferential 
procedure of choice (as the sampling distribution of the means will be even more 
"well-behaved" than the data distribution).

It is also interesting to observe a few things from the summary table. First, 
we observed negative values of `phi.j` in the majority of instances tested, 
which suggests an advantage of the MOEA/D-DE over the original MOEA/D (remember, 
_smaller = better_ for the quality indicator used). The MOEA/D-DE seems to 
require smaller sample sizes in most instances, suggesting a smaller variance of performance, which is also a desirable feature. Also, in three instances 
(UF\_5\_28, UF\_3\_29 and UF_\5_\17) the maximum number of runs/instance 
(`nmax = 200` in the `run_experiment()` call) was not enough to reduce the 
standard error (our "measurement error" on the values of `phi.j`) below the 
predefined threshold of $0.05$. There is no reason to worry in this particular 
case, however, since the resulting standard errors were not particularly high, 
and therefore their effect on the test power (resulting from the increased 
uncertainty in the estimation of these particular `phi.j` values) will be insignificant.

Since our observations `phi.j`  already express paired differences per instance, we can compare the two algorithms using a simple, one-sample t.test:

```{r}
t.test(my.results$data.summary$phi.j)
```

which indicates a statistically significant advantage of the MOEA/D-DE over the 
original MOEA/D ($p = 2.90\times10^{-6}, df=33$), with estimated mean 
IGD gains of $37.94\%$ ($CI_{0.95} = [24.22\%,51.66\%]$) over the original 
MOEA/D for instances belonging to the problem class of interest.

These results could also be used to motivate further analyses. For instance, 
we can observe in the summary table that the only two cases for which the 
MOEA/D was substantially better than the MOEA/D-DE were for different 
dimensions of problem UF\_5, which could suggest that some specific 
feature of this problem jeopardizes the latter algorithm's search mechanism. 
This could motivate research on which particular aspect of this problem results 
in this loss of performance, and on how to improve the MOEA/D-DE.

Finally, the full data of the experiment is contained in other fields of 
the output list `my.results`, and the user is encouraged to explore these. 
As an example, we can generate box plots and confidence intervals on the mean 
performance of each algorithm on each sampled instance, which could 
inspire new questions for the researcher.

```{r,fig.width=7,fig.height=7}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggridges))

# Adjust instance names for plotting
mydata <- my.results$data.raw
mydata$Instance <- gsub(pattern = "UF\\_", replacement = "UF", mydata$Instance)
mydata$Instance <- gsub(pattern = "\\_", replacement = " (", mydata$Instance)
mydata$Instance <- sapply(mydata$Instance, FUN = function(x){paste0(x, ")")})

ggplot2::ggplot(mydata, 
                aes(x = Observation, y = Instance, fill = Algorithm)) + 
  ggridges::geom_density_ridges(alpha = 0.7) + 
  ggplot2::ggtitle("Estimated IGD distribution", 
                   subtitle = "for each algorithm on each instance") + 
  ggplot2::theme(legend.position = "bottom")
```

```{r, fig.width=7}
# Calculate confidence intervals for each instance
algos <- unique(mydata$Algorithm)
ninstances <- length(my.results$instances.sampled)
CIs <- data.frame(instance  = rep(unique(mydata$Instance, times = 2)),
                  algorithm = rep(algos, each = ninstances),
                  x.est     = 0, CI.l = 0, CI.u = 0)

for (i in 1:ninstances){
  tmpdata <- mydata %>% 
    filter(Instance == unique(mydata$Instance)[i])
  myt1 <- t.test(tmpdata$Observation[tmpdata$Algorithm == algos[1]])
  myt2 <- t.test(tmpdata$Observation[tmpdata$Algorithm == algos[2]])
  
  CIs[i,3:5] <- c(myt1$estimate, as.numeric(myt1$conf.int))
  CIs[i + ninstances,3:5] <- c(myt2$estimate, as.numeric(myt2$conf.int))
}

# Plot individual confidence intervals for each instance
myplot <- ggplot2::ggplot(CIs, aes(x = instance, 
                                   y = x.est, ymin = CI.l, ymax = CI.u,
                                   group = algorithm, colour = algorithm,
                                   fill = algorithm))
myplot + 
  ggplot2::geom_pointrange(position = position_dodge(width = 0.5), alpha = 0.7) + 
  ggplot2::xlab("Instance") + ggplot2::ylab("IGD") + 
  ggplot2::ggtitle("Estimated mean IGD", 
                   subtitle = "for each algorithm on each instance") + 
  ggplot2::theme(legend.position = "bottom",
                 axis.text.x = element_text(angle = 55, hjust = 1, size = 6))
```
